Negative Types

### 2014-05-09 negatives.tex
story

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi$: Operational and Denotational Semantics}

Present $\Pi$ with an interpreter specifying the operational semantics. For a
model or denotational semantics, the minimum requirement is that every type
maps to some kind of space; a value $v : t$ maps to an element of that space;
and a combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
map a type of size $n$ to a finite set of $n$ elements; a value $v : t$ maps
to an element in the set denoting $t$; a combinator maps to a permutation on
that set. One could also build other models, e.g. vector spaces but that
model is fine. Perhaps show correspondence between operational and
denotational semantics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{-}$: Operational Semantics}

Extend $\Pi$ with negative types and $\eta$ and $\epsilon$. Present the
operational semantics that uses two interpreters forwards and backwards.

For the model or denotational semantics, we need to map every type to some
space. If type $t$ maps to space $S$, we need to have an operation that maps
that space to another space that acts like ``negative'' $S$ so that we can
satisfy for example that $S-S=0$. Using the cardinality of the set like we
did for $\Pi$ does not directly work (footnote on papers that describe sets
with negative cardinality). Another idea is to generalize spaces to have two
components (a producer and a consumer; a left player and a right player) and
to describe the space using the relative ``strength'' of the two
components. (This is the idea of the Int or G construction but that doesn't
directly work if we have two monoidal structures.) A natural and beautiful
theory is that of Conway games which model two player games: left and
right. In that setting we can talk about games in which the left player has
an $n$-move advantage (summarized by the number $n$) and games in which the
right player has an $n$-move advantage (summarized by the number $-n$ which
denotes that the left player has an $n$-move disadvantage). It is therefore
possible to map a type of size $n$ where $n$ is an integer to the class
Conway games where the left player has an $n$-move (dis-)advantage; the
negative of a space $S$ is the space in which all the games are flipped with
left and right players exchanging positions; a value $v : t$ maps to a Conway
number in the set denoting $t$; all combinators preserve the size of the type
and hence will be the identity. Note that the space corresponds to type $0$
is not empty in this denotation: it consists of all the games in which the
left player has a $0$-move advantage over the the right player and
symmetrically, i.e., the set of games which are always lost by the first
player to move.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conway Numbers}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correspondence}

Show correspondence between operational and denotational semantics for
$\Pi^{-}$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\textsc{LET}: Operational Semantics}

We introduce a toy functional language. Separate the first-order from the
higher-order component.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{o}$: Operational and Denotational Semantics}

Present $\Pi^{o}$ with an interpreter specifying the operational
semantics. This is $\Pi$ with recursive types and a trace operator. (It might
be enough to have just one recursive type $\mathbb{N}$.) We know
\emph{first-order} \textsc{LET} can be translated to this language: we cannot
translate higher-order functions. That requires compact closure but we live a
plain symmetric bimonoidal category.

For a concrete model or denotational semantics that instantiates the
categorical framework, the minimum requirement is that every type maps to
some kind of space; a value $v : t$ maps to an element of that space; and a
combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
map a type of size $n$ to a set of $n$ elements (including the infinite set
of $\omega$ elements; a value $v : t$ maps to an element in the set denoting
$t$; a combinator maps to a permutation on that set. One could also build
other models, e.g. vector spaces (infinite dimensional however) but that
model is fine. Perhaps show the details of the model and the correspondence
between operational and denotational semantics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{o-}$: Operational Semantics}

Extend $\Pi^{o}$ with negative types and $\eta$ and $\epsilon$. Present the
operational semantics that uses two interpreters forwards and backwards. Do
we have such a semantics that we like?

The categorical model is now supposed to be a ring category. This is
complicated so we will work directly with a concrete model that gives us more
operational intuition. We need to map every type to some space. If type $t$
maps to space $S$, we need to have an operation that maps that space to
another space that acts like ``negative'' $S$ so that we can satisfy for
example that $S-S=0$. Using the cardinality of the set like we did for
$\Pi^{o}$ does not directly work (footnote on papers that describe sets with
negative cardinality). Another idea is to generalize spaces to have two
components (a producer and a consumer; a left player and a right player) and
to describe the space using the relative ``strength'' of the two
components. (This is the idea of the Int or G construction but that doesn't
directly work if we have two monoidal structures and needs a complicated
completion to ring categories.) A natural and beautiful theory is that of
Conway games which model two player games: left and right. In that setting we
can talk about games in which the left player has an $n$-move advantage
(summarized by the number $n$) and games in which the right player has an
$n$-move advantage (summarized by the number $-n$ which denotes that the left
player has an $n$-move disadvantage). It is therefore possible to map a type
of size $n$ where $n$ is an integer to the class Conway games where the left
player has an $n$-move (dis-)advantage; the negative of a space $S$ is the
space in which all the games are flipped with left and right players
exchanging positions; following common terminology, we call the target of the
type an \emph{arena}; a value $v : t$ maps to a \emph{strategy} for the left
player in the arena for type $t$. Combinators transform strategies. Note that
the space corresponds to type $0$ is an arena with no possible moves.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conway Numbers: Arenas}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{o-}$: Denotational Semantics}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{o-}$: Correspondence}

Show correspondence between operational and denotational semantics for
$\Pi^{o-}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Back to $\textsc{LET}$}

Now work out the full translation from \textsc{LET} to $\Pi^{o-}$.
